---
title: "README"
output: github_document
date: "2023-08-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, echo = T, message = F)
```

## Fair Housing Map

In process documentation for CTCAC/HCD Opportunity and Neighborhood Change Map package, temporarily called the "Fair Housing Map" until directed further by CTCAC and HCD. Please see the [draft methodology](https://docs.google.com/document/d/1_wk_B9l0ZVSAr6bSoVROc7xuwTteyJxc/edit?usp=sharing&ouid=106736372141423442162&rtpof=true&sd=true) for details on the mapping approach.

### Installation

1.  Clone repository

2.  Download data-raw directory and extract into root directory: <https://berkeley.app.box.com/folder/224592265510>

3.  Open package using "fairHousingMap.Rproj" file

4.  Load package and install dependencies:

```{r, results='hide', message=FALSE, warning=FALSE}
devtools::install_dev_deps()
```

```{r results='hide', message=FALSE, warning=FALSE}
devtools::load_all()
```

### Final Dataset

The final dataset can be loaded by calling `data(final_2024)`. All variables are documented and the data dictionary can be loaded using `?final_opp_data`.

### Census API

1.  Register for Census API key [here](https://api.census.gov/data/key_signup.html).

2.  Install the key in the tidycensus package. See `?tidycensus::census_api_key()` for details

### Directory Structure

-   *data* - .Rda files of census polygon geographies.

    -   *intermediate* - intermediate data files.

-   *analysis* - .Rmd files of internal reports, exploratory analysis, etc.

-   *output* - Published datasets

-   *README* - project root documentation

-   *.gitignore* - instructions for git to ignore files

-   *R* - R package functions, executes all code that goes into published tables and maps

-   *map* - javascript webmaps

-   *tests* - testthat tests to ensure package is working as expected

### TODO

-   Develop testthat test suite (in progress as of 9/4/2023).
-   The legacy packages maptools, rgdal, and rgeos, underpinning the sp package, which is a dependency, will retire in October 2023. Replace functions that depend on sp with sf package equivalents.

### Running the package

Generating the final opportunity designations requires calling a sequence of functions in a specific order. Only functions which write intermediate files are necessary to generate a new final output.

#### **01_census_geography.R**

Contains `read_tract_centers()`, `read_block_group_centers()`, and `read_block_centers()`, functions for reading population weighted centroids. The centroid-reading functions are called within other subsequent scripts, and do not themselves produce any intermediate files.

#### **02_create_rural_shapefile.R**

Contains `shape_rural()` , which imports the USDA shapefile for areas ineligible for rural development housing programs, then shrinks to only California data and subtracts them from the complete California shapefile to get a rural areas shapefile. `shape_rural(write = T)`generates the first necessary intermediate file, rural_shapefile.shp, which must exist in the intermediate directory to proceed.

```{r, results='hide', message=FALSE, warning=FALSE}
shape_rural(write = F) # write = T will generate a new file
```

```{r, results='hide', message=FALSE, warning=FALSE}
mapview::mapview(shape_rural(write = F)) # to map the shapefile
```

#### **03_create_region_designation.R**

Contains `create_regions()`, which evaluates which regions that each county belongs to, then uses `rural_overlay()` to pinpoint rural tracts. `rural_overlay()` merges block centers with the rural_shapefile, and classifies the population of any block with its centroid inside the rural shapefile as rural. It then collapses to tract level, and any tract with over 50 percent population as rural is classified as "Rural Areas." `create_regions(write = T)` generates the required intermediate file, tract_county_region.csv.

```{r, results='hide', message=FALSE, warning=FALSE}
create_regions(read = T) # reads the existing file, set to False to generate new
```

#### 04_census_data.R

Contains `all_census_data`() which downloads all relevant decennial and ACS data from Census API using tidycensus, then derives percentages and calculates margins of error for derived variables. Both `all_census_data(geo = 'tract', write = T)` and `all_census_data(geo = 'bg', write = T)` are necessary to generate intermediate tract- and block group-level data.

```{r, results='hide', message=FALSE, warning=FALSE}
all_census_data(geo = 'tract', read = T, write = F) # for tracts
```

```{r, results='hide', message=FALSE, warning=FALSE}
all_census_data(geo = 'bg', read = T, write = F) # for block groups
```

#### 05_education_data.R

Contains `read_educ_pov()` which imports relevant math/reading scores and share of students receiving free or reduced priced meals (frpm). `read_educ_pov()` is called in subsequent scripts and does not produce any intermediate files. Reading/math data is weighted by total enrollment of schools that return 4th-grade test scores. FRPM data is limited to and weighted by enrollment in schools that serve elementary school students. `graduation_rates()` imports and writes cohort-weighted high school graduate rates to the intermediate directory.

```{r, results='hide', message=FALSE, warning=FALSE}
graduation_rates(write = F) # set to True to generate new data
```

#### 06_education_distance.R

Contains `school_distances()` which finds school distance to tract or block group centroids, and averages the reading and math scores, frpm, and graduation rates of the three nearest schools.

```{r, results='hide', message=FALSE, warning=FALSE}
school_distances(read = T, write = F) # set read to False, write to True to generate new data
```

#### 07_enviroscreen.R

Contains `xwalk_ces()` which imports CalEnviroScreen 4.0 site-based measurements (cleanup sites, hazardous waste, groundwater threats, solid waste) from the final 2023 TCAC file, creates a binary score for tracts in the bottom 5% of site-based hazards within regions, then crosswalks to 2020 tracts using an overlay method of \>= 5% of intersecting land area.

```{r, results='hide', message=FALSE, warning=FALSE}
xwalk_ces(read = T, write = F)
```

#### 08_neighborhood_change.R

Contains `read_neighborhood_change()` which imports neighborhood change data generated by CHPC. CHPC's scripts are located in R/neighborhood_change/ and identify tracts that have experienced both long-term (since 2000) and recent (since 2013) racial/ethnic and economic change. `read_neighborhood_change()` is called in the subsequent script.

#### 09_final_data.R

Contains `final_opp()` which loads and combines all intermediate files of economic, education, and environmental indicators, creates the final opportunity scores and designations. `read_neighborhood_change()` is called and joined to the final data frame.

```{r, results='hide', message=FALSE, warning=FALSE}
final_opp()
```

```{r, results='hide', message=FALSE, warning=FALSE}
mapview::mapview(final_opp(as_geo = T), zcol = 'oppcat') # to map the output
```
